The weight of the mountains: Constraints on tectonic stress, friction, and fluid pressure in the 2008 Wenchuan earthquake from estimates of topographic loading

### 2015-02-13 revisions/rebuttal2.md

>2 The fault geometry needs to be identified early in the manuscript. Right
now, the manuscript does not describe the fault dip anywhere, and this will
cause readers unnecessary difficulty.  These difficulties are particularly
apparent in section 3.

As described below, we have added descriptions of the faults in the text,
and made some figure annotations that we believe will be helpful. We state
here, as below, that the fault geometry is variable within each slip model
and between slip models, such that we cannot state a fault geometry as a
simple strike and dip.

>3 The manuscript confuses (or fails to clearly distinguish between) tractions
and stresses in several places. My experience is that this is common in the
Earth science community. It might be useful in an introductory section to state
that tractions are vectors that act on a particular plane, where as the stress
at a point is a tensor quantity.

We believe that most or all of our readers will know the difference
between tractions and stresses.  Stresses on faults are commonly
referred to as 'shear stress' and 'normal stress', instead of 'shear
traction' and 'normal traction'. One of us (EAH) has used both
terminologies in past papers. We also note that the influential paper
by Jim Rice ("Spatio-temporal complexities of slip on a fault", 1993,
JGR) uses "shear stress" for the stress acting on the fault, a
terminology that has become common in the "Rice school" of earthquake
dynamics.  Whether we are referring to tensors, vectors, or scalars
should be apparent and cause no confusion.

@@

> 5 The notation for stresses is unconventional, and that makes the manuscript
unnecessarily difficult to read. A more standard notation would be helpful.

We do not agree that it is quite so unconventional, and we have our reasons
for using the notation we chose; we have explained this below, and added some
discussion of this in the manuscript.

> 6 I think it would be better to describe certain distributions or functions
as posterior distributions or posterior functions rather than as “posteriors”
to avoid unintended reactions from some readers.

The term 'posteriors' has been well established in the literature for
decades and is unlikely to cause offense. Furthermore, the posteriors
we present in this paper are not function or distributions, but samples
of an unknown posterior PDF. We believe that our description of the
sampling methodology in section 4.2 makes this clear. We also note
that one of us was quite confused as what the the unintended reaction
from the term "posterior" might be.

@@

> 7 The figures and the text need to be better coordinated so that readers can
easily verify points made in the text when the figures are referred to. Several
of the figures can be modified to make them more useful to the reader. These
modifications should not be time-consuming to make.

We have made some minor modifications to figures to address this issue.

@@

> 8 A key modeling assumption seems to be that fault slip is not affected by
stress changes accompanying the fault rupture, but this assumption is not
stated. I think it should be.

We agree that this is important to state and did not realize we had
not in this manuscript. As stated, our model is  predicated
upon the Wallace-Bott assumption that fault slip occurs in the
direction of maximum shear stress, implicitly in that assumption is
that coseismic slip rake reflects accumulated shear stress on the
fault prior to earthquake nucleation, and that dynamic stresses do not
result in a significant re-orientation of coseismic slip. To make this
more clear, we have slightly modified the relevant sentence in the
abstract to:

"We estimate the tectonic stress needed to overcome topographic and
lithostatic stresses by assuming that the direction of maximum shear
accumulated on the faults is roughly collinear with the inferred
coseismic slip."

And in section 4.2, where we state the Wallace-Bott assumption, we
added "prior to the initiation of the earthquake" to make this clear.

@@

### 1. Introduction

**C1.1**

> The only omission in literature coverage I am aware of is the work by
McGovern on topographic stresses associated with volcanoes. I think some of
McGovern’s papers from the 1990’s should be cited.

We have looked over these references, which are largely about volcanic
deformation on other planets, and have not found a good place to cite
them, without making a separate paragraph for discussions of
topographic stresses in volcanoes. This is far from our topic (the
interaction of topographic stresses and tectonics), and we did not
rely on these papers while developing the work presented in this
paper, so we don't feel it necessary to cite them.

@@

We have added a paragraph and a half of description of the faults to the
introduction in the previous round of revisions.

**C1.3**

> (40-41) Referring to the Qur’an (or any other religious text) as one might
refer to a peer- reviewed scientific article is highly inappropriate. This
reference must be eliminated.

This reference has been eliminated, although we note that we feel it
 is unfortunate that these sorts of walls exist in scholarship. The
 history of ideas shouldn't be segregated so. We felt that the
 quotation was appropriate for the topic of this manuscript, and we
 did refer back to it later in the text. We are not sure that if we
 had chosen an opening quotation from a non-religious text, say James
 Joyce's Finnegans Wake, whether there would have been such a visceral
 reaction.  Finally, we did not refer to it 'as one might a
 peer-reviewed scientific article'. That all being said, we have
 removed the quotation.

@@

> This section came across well. I do think a description of the mechanics
behind equations 5 and 6 would be helpful though.

**C2.2**

> (169-170) The authors should be a bit more clear about some of the
limitations of the method of Liu and Zoback [1992]. That method cannot be used
to accurately calculate the topographically-induced stresses at shallow depths.
I think the method of Liu and Zoback is fine to use for the problem at hand,
but the method is not exact, and it does have its limitations. Those
limitations should be pointed out.

We change the sentence introducing the methods to "They show that the
topographic stress tensor field *beneath (but not within) topography*
can be determined by a convolution of topographic loading functions
with Green's functions describing the stresses in an elastic halfspace
due to a point load at the surface." (changes in
italics). Additionally, in Section 2.2 on the numerical implementation
of the methods, we mention that our calculations only apply to depths
below 851 m below sea level, which we take as our reference surface of
the half-space. We also now specifically state that z=0 corresponds to
mean sea level in our description of these calculations in Section
2.1.

@@

We have added a reference to Figure 4b and c.

**C3.8**

> (242) Where are the locations being referred to here? What are their x,y,z
coordinates (or latitude, longitude, and depth)?

The Pengguan fault is shown on the maps (Figures 1, 3, 9). The flats in the
Beichuan fault are at the base of the southwestern part of the fault, as is
evident in Figure 4, which shows some of the fault models. We think that giving
'x,y,z' or other spatial coordinates would be quite unhelpful (particularly for
curved surfaces tens of kilometers in breadth), so we will not do this.  If
readers need to know this information, they can look at the .csv files on
GitHub or consult the sources for these published slip models.

@@

**C3.9a**

> (243) What is the dip of the fault in question?

This varies in the different fault models considered. All of the
coseismic slip models we consider in this paper were developed by
other researchers (with the exception of Feng et al. (2010), of while
one of use is a co-author) and have been fully described in
peer-reviewed published papers. If the reader wants to know these
specifics, he or she is welcome to read the papers that have produced
the fault models we use.

@@

>$\tau_s$ = strike-slip traction on a fault

>$\tau_d$ = dip-slip traction on a fault

>This is much more in keeping with standard usage and would be easier for
(most) readers to follow I think.

We use $\tau$ for shear stress (tractions), which is quite standard
and we define our usage in the paper.  We also use $\sigma_n$ as
normal stress (traction) which is also standard and defined in the
paper. We use *M* to denote the topographic stress tensor or tensor
field, and it is clearly defined as such. Capital letters are
commonly, although not exclusively, used to denote tensors and tensor
fields, in a variety of sources, spanning structural geology, fluid
dynamics, mathematics, and engineering disciplines.

@@

We disagree that a figure is necessary, or particularly helpful, to verify a
claim that vertical stresses resolve more as normal stress than shear stress
on a plane with low dip. This should be apparent to readers who have taken a
course in Structural Geology or who have a continuum mechanics background,
which surely includes the vast majority of our readers. Readers without this
background will unfortunately have a quite difficult time with the paper, but
they are not our intended audience.

**C3.11**

> (249-251) I don't see where to look to verify the nature of the slip and the
tractions on Figure 4.  The blue background color of the slip magnitude figure
(upper left) looks very uniform, so I can't tell where the slip is high; also
note that the contour lines are not labeled with magnitudes. In the normal
stress plot in the upper right the red background color is very uniform, so
I can't tell where the normal traction is high (and the contour lines are not
labeled). This means I don't know how to compare the plots.

The colors shown are not simple 'background colors' but correspond to
the magnitudes of the stress component in question, this is stated in
the figure caption and the associated color bars in the figure. Where
they are uniform, stresses are uniform. Labeling the contours would
add unnecessary	 clutter to the figures, making them harder to
read. The relevant comparisons (for example, between high normal
stress and low slip) are still quite evident to us, both on the screen
and in print, and other viewers have not had problems with this.

@@

### 4. Calculations of tectonic stress, fault friction and pore fluid pressure

**C4.1**

> This section reads fairly well, although a fuller description of the Bayesian
statistics would have helped me.

We believe that we have explained the mathematics involved in our
calculations (Bayes' rule, the misfit equation, and the likelihood
equation, which are equations 10-12 in the original submission and
equations 11-13 in the revised submission). These are uncomplicated
expressions, and we believe that they fully describe our estimation
strategy within a Bayesian sampling methodology.  We have tried to
define and explain each of the terms, but we do not feel that a more
basic and general introduction to Bayesian methods is appropriate in
this paper. We cited two excellent texts [Tarantola, 2005; Mosegaard
and Tarantola, 1995] that explain Bayesian estimation in more detail.

@@

**C4.2**

> (256-257) The statement here should also account for cohesion, although the
frictional resistance could be much larger than any intrinsic cohesion

This is true; we have added "and assuming that cohesion is negligible"
to the sentence describing the Mohr-Coulomb failure envelope.
***NOTE: I will remove "rock strength" from the main text, since you are talking about a failure on a pre-existing fault, it is misleading, let's just leave as "cohesion".***

@@

Yes, this is what we meant.  We have replaced 'fault' with 'earthquake' in this
sentence.

**C4.4a**

> (265) The phrase "posteriors of T" could have unanticipated (and undesirable)
reactions from some readers. I suggest using the alternative phrase "these
functions".

The use of the word 'posterior' in statistics is well-established and
is fairly commonly used in geophysical research papers that utilize
Bayesian estimation or inferrence. Additionally, as described in
section 4.2, our sampling estimation strategy results in samples of
some unknown posterior probability density function, this is the
solution to the inverse problem in a Bayesian sense, and so refering
to these collections of posterior samples as "functions" would be
misleading.

@@

We have fixed these problems as part of the previous revisions.

**C4.6**

> (276-287) Owing to my poor background in Bayesian statistics, I cannot
comment knowledgeably on the technical aspects of this paragraph except for one
point.  The authors should clarify whether the are evaluating the direction of
slip based on the conditions before the fault has begun to slip or as the fault
slips and the stress field around the slipped patch is evolving.

The direction of slip is given by the coseismic slip models, which are
the net slip from the earthquake; this is described in Section 4.2.
We think the reviewer is addressing whether under the Wallace-Bott
assumption the final coseismic slip rake is parallel to the direction
of maximum shear stress accumulated on the fault prior to slip
initiation, and that dynamic stresses do not affect the coseismic slip
rake. We address this point in comment 8 above.

@@

The placement of the key points of a section to be in the last sentence of that
section is a stylistic preference, and not one that we share (why not put the
important information towards the top of the paragraphs rather than at the
end?). There is not really a 'key point' here per se; this is a description of
the results. The important results are discussed as such.

**C5.3**

> (328-330) After this sentence, say what the significance of the peak in
T'(min) at 0.2 is, as was done in lines 326-328 for T'(max).

We don't view the value of the minimum tectonic stress to be as
significant as the value of the maximum stress, and there is no simple
statement that can summarize the importance of this value (i.e. it
does not directly affect much).  We do however discuss the results
more in the discussion.

@@

We think the most likely values in our results are the most likely. When we
stated the conditions that were unlikely, we were explaining why they
were not well represented in the posteriors.

**C5.6**

> (341-345) I can't see a peak in T'(Max) at ~0.6 in any of the plots on Fig.
8. Some additional explanation is needed, at least for me.

The highest values of the $T'_{\max}$ marginal histograms at around
0.6 are evident. This 'peak' (not our wording) is not particularly
sharp, but it is clearly there; there is a steep decrease in
probability to the left of the mode and a more gradual decrease to the
right of the mode. As stated in the response letter to the previous
revision, the likelihood at 2.5 (the upper boundary of $T'$ is 60% of
the mode at 0.6.  We have amended this sentence to read '0.6-0.8' so
it is clear that we do not mean a narrow range of high likelihoods
around a single value.

@@

We have changed the sentence to read: "However, $p_{\mathrm{J}}(\mu | D)$ has
a lower relative likelihood on the high-$\mu$ tail (Figure 8d) compared to
$p(\mu | D)$ of the constituent slip model results (Figure 7)."

> “This lower likelihood of m in the joint posterior is likely bec[au]se it is
the average μ of all slip models. On the other hand, it is not similarly sparse
on the low-μ side, suggesting that low values for μ are more robust.”

> <u>How low is "low"? Can the authors be quantitative here?</u>

We added a reference to the figure (mentioned in the last response).
We would prefer that readers look at the distribution itself, as this
is the quantitative solution of the inverse problem in a Bayesian
sense, rather than a number we throw out.

@@

As we state in the paper, these calculations apply to each point in the
coseismic slip model, so the optimal faults would be collocated with the
points of the slip model. Because we treat each point independently and
the comparisons are of local stresses, it doesn't make a difference that
with different orientations, the optimal faults wouldn't align into a single
plane.

**C6.18**

> (434-479) The conclusions of this paragraph, that the Beichuan fault is more
likely to slip than an optimally-oriented fault, will strike most readers as
impossible. How could the Beichuan fault be more optimally oriented for slip
than a fault that has the optimal orientation for slip? The "best" (in terms of
slip probability) that could be hoped for is that the Beichuan fault has the
orientation (and position) of the optimally-oriented fault, and therefore is
the optimally-oriented fault. How is it possible that slip on the Beichuan
fault is favored by a factor of as much as 1000 (the anti-log of 3) over the
optimally-oriented fault? I think the answer lies in the definition of what
constitutes the optimally-oriented fault. This needs to be more clearly and
completely explained to avoid confusing readers.

As described in the manuscript, the 'optimally-oriented fault' has a
canonical value of friction of 0.6 (lines 466-469 of the submitted
manuscript). In contrast, our model results indicate that the faults
that are thought to have ruptured in this event have a much lower
coefficient of static friction.  As is shown in Table 2, the
coefficient of friction on the fault has a strong influence on the
favorability of slip on the Wenchuan faults vs. optimally oriented
faults of the stress parameters studied.  This is stated repeatedly
throughout this section, so we do not believe that we are unclear
about this point.

@@

Additionally, when we did the log transformation of the y-axis of Figure 10,
we used the base-*e* log, not base-10; the anti-log of 3 in this case is ~20,
which is far more reasonable value than 1000. We have updated the text and
figure captions so that 'log' now reads 'ln' to avoid this confusion.

Nonetheless, we have reviewed our code and found a slight error in how
the optimal slip plane was calculated; essentially, the algorithm
assumed that dip slip faulting would be favored (where the optimal
plane strikes perpendicular to the trend of $\sigma_1$) but produces
an incorrect orientation otherwise.  We have fixed the bug, which has
changed these results slightly (the bug only affected these
calculations; that particular function was not used in the rest of the
study). The new Figure 10 shows that more of the data plot in the
favorability region for optimal faulting than previously; however,
most of the data (with low friction values) still plot in the
favorability region for the Wenchuan faults.

@@

This is a LaTeX problem which we couldn't easily fix and won't be
an issue with the final version.

**CAp2**

> Equations (A4)-(A6) match the solutions in Mal and Singh (1991), but
equations (A1)-(A2) do not match and have errors. Owing to the different
elastic coefficients Mal and Singh use, I can’t tell if equation (A3) is
correct or not. Also note that G must have dimensions of force/area, which
equations (A4)-(A6) do, but the lead terms in square brackets in equations (A1)
and (A2) do not. Aside from those differences, the elastic coefficients need to
be defined. Similarly, equations (A10) and (A12) match the solutions in Mal and
Singh (1991), but I can’t tell whether equations (A7)-(A9) do because of the
different elastic coefficients. Since the authors did not derive these
equations, they could simply cite a reference and they would not need to
include an Appendix, although they should check to make sure the errors in
equations (A1) and (A2) did not spill over to their codes.

The equations have been amended (a *z* was missing in the numerator of
the first term in the brackets). The mistake is not found in the code.
This fixes the unit problem mentioned.  We reproduce the equations
from Liu and Zoback [1992] here for completeness. Since we define
these equations using different notation, we beleive that if we just
referred to Liu and Zoback [1992] there might be some confusion for
readers trying to reproduce our results. We thank the reviewer for
catching the typo in equations (A1) and (A2).

@@

Yes, $\sigma^B$ should be $M^B$, and $F_h,y$ instead of $F_h,yz$.

**CAp4**

> In equation (5) why are all three terms on the right side of the equal sign
in the first row multiplied by dh/dx? I can see why the ρgh term is, because if
the surface has a slope, then the slope will impart a horizontal shear stress.
I don’t understand why Mxx and Txx are multiplied by dh/dx though, because Mxx
and Txx already act horizontally. Similarly, why are Mxy and Txy multiplied by
dh/dy in the second row of the equation? Analogous comments apply to equation
(6). An explanation of this would be helpful.

Liu and Zoback [1992] devote some attention to the horizontal loading
function (primarily, see the "Effect of Topography on Tectonic Stress"
Section in their paper). While we do state in the beginnign of the
section that we use methods derived by Liu and Zoback [1992], we
realized that we should also cite their paper where we give the
horizontal loading function, and we have edited the sentence equation
(5) to state that these loading functions are derived by Liu and
Zoback [1992].

Yes, Mxx, Txx, Mxy, and Txy act horizontally, but they represent
normal tractions in horizontal directions on vertical planes, and not
tractions on the free surface. Mxx, etc, aboce the reference level,
here taken to be mean sea level, only excite shear traction on the
free surface (warranting their inclusion into the horizontal loading
function) if topography varies. We think of this as due to the fact
that you are changing the total size of the vertical planes that
tractions resulting from Mxx, etc, are acting upon. This change in
horizontal normal tration results in a shear stress on the reference
surface and stresses are propagated into the half-space via the
Cerruti solutions.
